---
title: "Advanced Macroeconomics 3"
author:
date: "24.1.2020"
output: html_document
---

## Problem set 1, Exercise 1.

```{r setup, include=FALSE}
# This package is needed for the filtering
pkgs <- c("mFilter", "dplyr", "data.table", "xtable", "DT")
lapply(pkgs, library, character.only=TRUE)

# Download data
setwd("Z:/Exercises")
data <- readxl::read_xlsx("Finland_data.xlsx")
data <- data[-c(1:2),]
```


GDP and consumption are defined as logarithmic time series objects. They are de-trended using two sided HP-filter. Lambda is set to 1600 according to standard practices with quartal time series data.

```{r variables, results=F}
y <- log(ts(data[,2], start = c(1975,1), end = c(2018,4), frequency = 4))
c <- log(ts(data[,3], start = c(1975,1), end = c(2018,4), frequency = 4))

y_hp <- hpfilter(y,freq=1600,type=c("lambda"),drift=FALSE)
c_hp <- hpfilter(c,freq=1600,type=c("lambda"),drift=FALSE)

y_trend <- y_hp$trend
c_trend <- c_hp$trend

y_detrended <- y_hp$cycle
c_detrended <- c_hp$cycle
```


###  Visualizing the results

```{r plots, echo=FALSE}
# Visualize the results
plot(y)
lines(y_trend, col=c("blue"))
title("GDP")
legend("topleft", legend=c("Real series", "HP-trend"), col=c("black", "blue"), lty=c(1), cex=1.2, box.lty=1)

plot(c)
lines(c_trend, col=c("blue"))
title("Consumption")
legend("topleft", legend=c("Real series", "HP-trend"), col=c("black", "blue"), lty=c(1), cex=1.2, box.lty=1)


```



## a)

Re-creating a section of the table in the slides for Finnish data.

```{r wrangle, echo=T}
## Calculate the values for the table
# Deviations
y_dev <- sd(y_detrended)
c_dev <- sd(c_detrended)

# Relative deviations
y_relative <- y_dev/y_dev
c_relative <- c_dev/y_dev

# First order autocorrelations
# First order autocorrelations
y_autocorr <- acf(y_detrended, lag.max = 1, plot = F)$acf[[2]]
c_autocorr <- acf(c_detrended, lag.max = 1, plot = F)$acf[[2]]

# Contemporaneous correlation with output
y_corr <- cor(y_detrended, y_detrended)[[1]]
c_corr <- cor(y_detrended, c_detrended)[[1]]

output <- cbind(c(y_dev, c_dev), c(y_relative, c_relative), c(y_autocorr, c_autocorr), c(y_corr, c_corr))
colnames(output) <- c("Standard deviation", "Relative standard deviation", "First order autocorrelation", "Contemporaneous correlation with output")
rownames(output) <- c("Y", "C")

```


### Results with two sided filtering

```{r table}
# Prepare the table
# print(xtable(output, digits=2), type="html")

datatable(output) %>% formatRound(columns=c(1:4), digits=2)

```




## b)

Estimate the size of the end point bias of HP-filter by comparing the previous results to those calculated form series, which were de-trended with one sided HP-filter.

```{r one sided}
# Custom function to conduct the analysis
one_sided_hp <- function(data, lambda){
    trend <- c()
    detrended <- c()
    
    for(i in 4:length(data)){
        x <- data[1:i]
        
        x_hp <- hpfilter(x, freq=lambda,type=c("lambda"),drift=FALSE)
        trend[[i]] <- x_hp$trend[[i]]
        detrended[[i]] <- x_hp$cycle[[i]]
    }
    
    trend <- ts(trend, start = c(1976,1), end = c(2018,4), frequency = 4)
    detrended <- ts(detrended, start = c(1976,1), end = c(2018,4), frequency = 4)

    return(list(trend=trend, detrended=detrended))
}



y_1s <- one_sided_hp(data=y, lambda=1600)
c_1s <- one_sided_hp(data=c, lambda=1600)

y_trend_1s <- na.omit(y_1s$trend)
c_trend_1s <-  na.omit(c_1s$trend)

y_detrended_1s <-  na.omit(y_1s$detrended)
c_detrended_1s <-  na.omit(c_1s$detrended)

```


### Visualizing the end point bias of the HP-filter in trends

```{r plot new, echo=F}
plot(y, lwd=1.5)
lines(y_trend, col=c("blue"))
lines(y_trend_1s, col=c("red"))
title("GDP")
legend("topleft", legend=c("Real series", "2 sided", "1 sided"), col=c("black", "blue", "red"), lty=c(1), cex=1.2, box.lty=1)

plot(c, lwd=1.5)
lines(c_trend, col=c("blue"))
lines(c_trend_1s, col=c("red"))
title("Consumption")
legend("topleft", legend=c("Real series", "2 sided", "1 sided"), col=c("black", "blue", "red"), lty=c(1), cex=1.2, box.lty=1)

# Visualize the end point bias of the HP-filter in cyclical components

plot(y_detrended, col=c("blue"))
lines(y_detrended_1s, col=c("red"))
title("GDP")
legend("topleft", legend=c("2 sided", "1 sided"), col=c("blue", "red"), lty=c(1), cex=1.2, box.lty=1)

plot(c_detrended, col=c("blue"))
lines(c_detrended_1s, col=c("red"))
title("Consumption")
legend("topleft", legend=c("2 sided", "1 sided"), col=c("blue", "red"), lty=c(1), cex=1.2, box.lty=1)

```


#### The series de-trended with one sided hp-filter exerts larger variances than those which were treated with the two sided version.

```{r wrangle again, echo=F}
## Calculate the values for the table
# Deviations
y_dev_s1 <- sd(y_detrended_1s)
c_dev_s1 <- sd(c_detrended_1s)

# Relative deviations
y_relative_s1 <- y_dev_s1/y_dev_s1
c_relative_s1 <- c_dev_s1/y_dev_s1

# First order autocorrelations
# First order autocorrelations
y_autocorr_s1 <- acf(y_detrended_1s, lag.max = 1, plot = F)$acf[[2]]
c_autocorr_s1 <- acf(c_detrended_1s, lag.max = 1, plot = F)$acf[[2]]

# Contemporaneous correlation with output
y_corr_s1 <- cor(y_detrended_1s, y_detrended_1s)[[1]]
c_corr_s1 <- cor(y_detrended_1s, c_detrended_1s)[[1]]

output_s1 <- cbind(c(y_dev_s1, c_dev_s1), c(y_relative_s1, c_relative_s1), c(y_autocorr_s1, c_autocorr_s1), c(y_corr_s1, c_corr_s1))
colnames(output_s1) <- c("Standard deviation", "Relative standard deviation", "First order autocorrelation", "Contemporaneous correlation with output")
output_comparison <- rbind(output, output_s1)
rownames(output_comparison) <- c("Y one sided", "C one sided", "Y two sided", "C two sided")

```



### Comparison of the results

```{r comparison, echo=F}
# Prepare the table
datatable(output_comparison) %>% formatRound(columns=c(1:4), digits=2)

```

#### Additional to higher variance, the one sided filtering also produces higher relative variance, contemporaneous correlation of consumption with output, and more persistent first order autocorrelation in both series.




